%
% DO NOT CHANGE THIS TEMPLATE!
%
% To allow combining the lecture notes from different scribes in a
% simple way all scribes must use the same header.
%
% If you think the template lack a package, then contact the lecturer,
% and we will probably extend the template with your favorite package.
%

\documentclass[11pt,a4paper,twoside]{article}
\usepackage{alg}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{algorithmic}

\begin{document}
\pagestyle{myheadings}
\thispagestyle{plain}

\newcommand{\lecture}[4] { %
\newcommand{\fdatum}{#1}%
\newcommand{\fnummer}{#2}%
\newcommand{\frubrik}{#3}%
\newcommand{\fskribent}{#4}%
\markboth{DD2458 -- Popup VT 2011}{\frubrik}%
\hrule
\begin{center}
\large \bf 
DD2458, Problem Solving and Programming Under Pressure

\vspace*{2mm}
\Large \bf 
Homework solutions week \fnummer: \frubrik
\end{center}
\small
\noindent Date: \fdatum
\\
Scribe(s): \fskribent
\\
\hrule
\vspace{5mm}
}


\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{definition}[theorem]{Definition}

% \lecture{date}{lecture number}{lecture headline}{lecturer}{scribe}
%
%
%

%%%%% Edit after this row. The above should not need editing. %%%%%

%%%%% Define your own stuff here. %%%%%

\newcommand{\zed}{\mathbb{Z}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\GF}[1]{\mathrm{GF}_{#1}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\lecture{2011-04-14}{9}{Number Theory}{Joel Bohman, Linus Wallgren, Oskar Werkelin Ahlin}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 1) FACTORIAL FACTORS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Factorial Factors}

In this problem, we want to find the maximum number of factors that can be
multiplied together to produce $N!$, given a number $N$.  The maximum number of
such factors is found in the prime factorization of the number. Finding such a
factorization for a number can easily be done using trial division. But in this
case, we want to find it for the factorial of the given number.  Let us define
a function $n(N)$ that denotes the number of prime factors in the factorization
of a number $N$. Then
\\\\
$n(N!) = n(2 \cdot 3 \cdot ... \cdot (N-1) \cdot (N)) = n(2) + n(3) + ... + n(N-1) + n(N)$
\\
\begin{algorithm}
    \caption{Sieve of Eratosthenes}
    \label{lst:primesieve}
    \begin{algorithmic}
        \STATE prime[] \COMMENT{A boolean list with size N, all set to \textbf{true}}
        \FOR{$i = 2 \to N$}
            \IF{$prime[i] = true$}
                \FOR{$j = 2 \to N/i$}
                    \STATE $prime[i \cdot j] = \textbf{false}$
                \ENDFOR
            \ENDIF
        \ENDFOR
    \end{algorithmic}
\end{algorithm}
            
\noindent{Use} the Sieve of Eratosthenes (Algorithm \ref{lst:primesieve}) to
generate all primes in the range, and save these in a list. To calculate
$n(x)$, we try dividing $x$ by the primes in our list. When we find such a
number $d$, we return $n(x/d) + n(d)$. The base case in the recursion is when
it is called with a prime number. In this case, return 1.  This can be checked
in constant time with the table used in the sieve.

% TODO correct?
%n(x):
%if x is prime:
%    return 1
%else:
%    for prime p in primes_list:
%        if p divides x:
%            return n(x/p) + 1

\begin{algorithm}
    \caption{$n(x)$}
    \label{lst:nx}
    \begin{algorithmic}
        \IF{$x$ is prime}
            \RETURN $1$
        \ENDIF
        \FOR{prime $p \le \sqrt{N}$}
            \IF{$p | x$}
                \RETURN $n(x/p) + 1$
            \ENDIF
        \ENDFOR
    \end{algorithmic}
\end{algorithm}
    
\noindent{Before} returning the number of factors in a number, we save it in a
table, to avoid doing the same job more than once. To get the final answer, we
just have to go through the range from $2$ to $10^6$
(which is the maximum value in the KATTIS problem), and add up the number of
factors, like this:

% TODO correct?
%count[i] = 0 for all i
%for (i = 2 to 10^6)
%    count[i] = count[i-1] + n(i)

\begin{algorithm}
    \caption{Count factors}
    \label{factorcount}
    \begin{algorithmic}
        \STATE $count[1] = 0$
        \FOR{$i = 2 \to 10^6$}
            \STATE $count[i] = count[i-1] + n(i)$
        \ENDFOR
    \end{algorithmic}
\end{algorithm}

\noindent{The} answer for a number $x$ is now found in $count[x]$. 
\subsection{Time complexity}
The sieve has time complexity $\mathcal{O}(N\log{\log{N}})$. The number of primes $\le \sqrt{N}$ is $\mathcal{O}(\sqrt{N})$ and thus reducing a number to its prime factors using trial division (as we do) is $\mathcal{O}(\sqrt{N})$. The precomputation step would generate a time complexity of $\mathcal{O}(\sum_{k=2}^{10^{6}}\sqrt{k})$. The practical time complexity is much better, as we keep values for the numbers in a table. For large numbers, the algorithm will execute in constant time. For large non-prime numbers, say $10^6$, the prime factors will quickly be detected, and the number will efficiently be reduced to a much smaller number, for which the probability that we have a precalculated value for it, is high.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 2) JACKPOT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Jackpot}

The problem statement asks us to find the jackpot periodicity $P$ of an entire
machine, given the jackpot periodicities $p_{1},p_{2},...,p_{n}$ for the
individual wheels in the machine. The answer is the smallest integer P such
that $p_{i}$ divides P, for all $i \in {1,2,...,n}$. If P $> 10^{9}$, we should
print "More than a billion.".
\\\\
\noindent{This} integer is found by computing the least common multiple of
integers $p_{1},p_{2},...,p_{n}$. LCM can be computed using an implementation
of GCD (algorithm \ref{lst:LCM}), in the following way:

\begin{algorithm}
    \caption{LCM - Least common multiplier}
    \label{lst:LCM}
    \begin{algorithmic}
        \REQUIRE Input: $a$, $b$
        \RETURN $ \frac{a}{gcd(a, b)} \cdot b$
    \end{algorithmic}
\end{algorithm}

When we want to calculate the least common multiplier of several numbers we can
use the fact that $LCM(a, b, c) = LCM(LCM(a, b), c)$. Because of this property
we can calculate the LCM of the current periodicity and the cumulative
periodicity of all previous wheels, and do so until either the comulative
periodicity is larger than $10^9$ or we have calculated the periodicity of the
entire machine.

\begin{algorithm}
    \caption{Jackpot}
    \label{lst:jackpot}
    \begin{algorithmic}
        \REQUIRE Input: wheels[] \COMMENT{Wheel periodicity}
        \STATE r = 1
        \FOR{each periodicity $p$ in $wheels$}
            \STATE $r = LCM(r, p)$
            \IF {$r > 10^9$}
                \PRINT More than a billion.
            \ENDIF
        \ENDFOR
    \end{algorithmic}
\end{algorithm}

\subsection{Time complexity}
The time complexity for finding the $LCM$ of two numbers $a,b$ is $\mathcal{O}(\log{max(a,b)})$. Total time complexity is $\mathcal{O}(k*log{p_i})$, where $p_i$ is the periodicity and $k$ the number of wheels.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 3) REPEATING DECIMALS *
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Repeating Decimals (*)}

In this problem, we want to find the repeating decimal cycle of a rational
number $\frac{p}{q}$, given integers $p$ and $q$. None of these will exceed
$3000$, and they will be positive. We want to print the number in its decimal
notation, placing the repeating decimals inside parentheses. We also want to
count the number of decimals inside the repeating cycle.
\\\\
\noindent{The} left-most part of the decimal mark is found in
$\lfloor{\frac{p}{q}}\rfloor$. From this division, we get a remainder. Let us call
the remainder $r$, and give it the value $p \mod q$. To get the decimal expansion,
we let $r = r \cdot 10$. Now the first digit after the decimal mark is found in
$\lfloor{\frac{r}{q}}\rfloor$. This procedure will repeat until we get the same
remainder that we got in a previous step. It is clear that when this happens, we
will get an infinite loop of the same cycle.
\\\\
\noindent{We} need an efficient way to store the remainders we had and where they
appeared. The maximum number of different remainders is limited to the maximum
possible value of $q$ (since the remainders are calculated $\mod q$). This means
that there is a maximum of $3000$ remainders for a number. Use an integer array to
map a remainder to its first occurrence in the decimal expansion, and let all unseen
remainders have value $-1$.  Let us also use the array $digit[]$ to save the actual
digits in the expansion.

% TODO
%init seen[] to -1
%r = p % q
%seen[r] = 0 // first remainder seen at index 0
%set i = 0
%while (true)
%    if (r == 0)
%        break
%    r *= 10
%    digit[i] = $\floor(r/q)$
%    r = r%q
%    if (seen[r] != -1)
%        // the cycle starts at seen[r] and ends at i
%        break
%    seen[r] = i+1
%    ++i

\begin{algorithm}
    \caption{Long division}
    \label{lst:Long division}
    \begin{algorithmic}
        \STATE Initialize $seen[]$ to $-1$
        \STATE $r = p \mod q$
        \STATE $seen[r] = 0$ \COMMENT{First remainder seen at index 0}
        \STATE $i = 0$
        \WHILE {$r \neq 0$}
            \STATE $r = r \cdot 10$
            \STATE $digit[i] = \lfloor\frac{r}{q}\rfloor$
            \STATE $r = r \mod q$
            \IF {$seen[r] \neq -1$}
                \STATE \COMMENT {The cycle starts at $seen[r]$ and ends at $i$}
                \STATE \em{break} % TODO Make bold
            \ENDIF
            \STATE $seen[r] = i + 1 $
            \STATE $ i = i + 1$
        \ENDWHILE
    \end{algorithmic}
\end{algorithm}

       
\noindent{If} the decimal expansion is finite (e.g. we get remainder zero in the
division), the repeating cycle has length one and contains a single zero.
Otherwise, we print the decimal expansion. In the KATTIS problem, we don't want
to print any more than $50$ digits. If the cycle ends after more than $50$
digits (from the decimal mark, not the start of the cycle) we print "..." and
end it with a right parenthesis. The number of decimals in the cycle is $1$ in
the case of $r==0$, otherwise it is found in the difference between the end and
starting point of the cycle, as seen in the pseudo code above.

\subsection{Time complexity}
Let $B$ be the number of digits before the cycle starts, and $C$ the cycle length. The time 
complexity for finding the cycle is $\mathcal{O}(B+C)$. As we are performing some slow 
operations (like modulus), the practical performance will be a bit slower than a standard 
linear algorithm.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 4) THE STERN-BROCOT NUMBER SYSTEM
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Stern-Brocot Number System}

The Stern-Brocot number system is an elegant way to express all fractions
$\frac{p}{q}$ where $p$ and $q$ are relatively prime. In this problem, we want
to find the path from the root of the tree to a fraction $\frac{p'}{q'}$, given
$p'$ and $q'$. 
\\\\
\noindent{We} call the current node $C$. We start our path in the root node
with $C = C_{p}/C_{q}, C_{p} = 1, C_{q} = 1$. Let us also call the target node
$T = T_{p}/T_{q}$. If $T > C$, we go right, otherwise we go left. When changing
the current node, we let $C = C + L$ if we went left, and $C = C + R$ if we
went right. $L$ and $R$ are pointers to the two adjacent nodes, initially set
to $\frac{0}{1}$ and $\frac{1}{0}$. Note that we change the $L$ and $R$ pointer
when switching nodes.

% TODO
%Parent_l = a/b
%Parent_r = c/d
%Child = (a+c)/(b+d)

\noindent{In} the algorithm we use a simple structure, fraction, which stores a
fraction as its numerator and denominator.

\begin{algorithm}
    \caption{Calculate path in the Stern Brocot number system}
    \label{lst:sternbrocot}
    \begin{algorithmic}
        \STATE fraction $left = \frac{0}{1}$
        \STATE fraction $right = \frac{1}{0}$
        \STATE fraction $current(left, right)$ \COMMENT{Compute a new node from its two parents, left and right}
        \WHILE{$current \neq target$}
            \IF {$target > current$}
                \STATE $left = current$
                \PRINT R
            \ELSE
                \STATE $right = current$
                \PRINT L
            \ENDIF
            \STATE current = fraction(left, right) \COMMENT{Generate a new node from the new parents}
        \ENDWHILE
    \end{algorithmic}
\end{algorithm}

\noindent{One} has to be careful when implementing the comparison operator
between two fractions.  Note that $a/b > c/d$ ($b,d \neq 0$) is the same as $a
\cdot d > c \cdot b$, which gives better precision.

As the algorithm simply walks down a binary tree the time complexity is
$O(\log(N))$, where $N$ is the number of nodes in the tree.

% TODO
%fraction left(0,1)
%fraction right(1,0)
%fraction current(left, right) // Compute the current node from the two parents, left and right
%    while (current != target)
%        if (target > current)
%            left = current
%            print “R”
%        else
%            right = current
%            print “L”
%        current = frac(left, right) // Generate a new node.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 5) RIEMANN VS MERTENS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Riemann vs. Mertens}

\noindent{The} naive solution is to simply use trial division to factor each
number from $1$ to $1000000$, and depending on the number of factors and if
there is two of one of them, assign the correct value. This would take far too
long time, so we need to do something smarter.

\noindent{Since} we need to calculate the Mobius function for all values in the
range we can use every prime number less than or equal to $\sqrt{1000000}$ with
algorithm \ref{lst:primesieve}, and square it, and mark that off as a number
that is not square free.

\noindent{The} next step is to calculate the number of factors in each number.
For each number we can simply use trial division to remove one factor and
recursively call the Mobius function for that value. As performance will be an
issue when doing this we can simply cache the results, as they will be used
more than once, and in order to get cleaner code we can add $-1$ to all primes
in the beginning, when we do the prime sieve to calculate them.

\begin{algorithm}
    \label{mobius}
    \caption{The mobius function}
    \begin{algorithmic}
        \REQUIRE Input: $N$
        \IF{$cache[N]$ is DEFINED}
            \RETURN $cache[N]$
        \ENDIF
        \FOR{prime $p \le N$}
            \IF {$p | N$}
                \STATE $cache[N] = - mobius(N/p)$
                \RETURN $cache[N]$
            \ENDIF
        \ENDFOR
    \end{algorithmic}
\end{algorithm}

\noindent{The} Mertens function has to be calculated in $O(N)$ time, simply
iterate through all numbers and keep track of the value from the last iteration
and add the Mobius value from this iteration. As each value in the Mertens
function is dependant on the previous ones, and we know we are going to have to
do a lookup more than once, we can simply calculate every value and save it.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 6) SEMI-PRIME H-NUMBERS *
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Semi-prime H-numbers (*)}

This problem asks us to count the number of H-semi-primes within a certain 
range. We do this in three steps.

\noindent{In} the first step we want to create a list of all H-prime numbers in
the sequence created by $4n + 1$ for any positive $n$. An H-prime is the
product of only the unit (which is $1$) and the H-prime itself. We begin by
setting all numbers in the range as H-primes and use a modified Sieve of
Eratosthenes to find all non-prime H-numbers and mark those in the list.

\begin{algorithm}
    \caption{H-prime sieve}
    \label{lst:hsieve}
    \begin{algorithmic}
        \STATE $i = 5$
        \WHILE{$i \le \sqrt{LIMIT}$}
            \IF{$i$ is H-prime}
                \STATE $j = i \cdot i$
                \WHILE{$j \le LIMIT$}
                    \STATE $j$ is not H-prime % TODO det här ser fel ut
                    \STATE $j = j + i$
                \ENDWHILE
            \ENDIF
            \STATE $i = i + 4$
        \ENDWHILE
    \end{algorithmic}
\end{algorithm}

\noindent{In} the next step we take an H-prime number from the previous list we
created. We then take another H-prime number which is the same or larger and
calculate the product of those. The product of those are a semi-prime H-number.
This way, we generate all possible semi-prime H-numbers.

\begin{algorithm}
    \caption{Calculate semi-prime H-numbers}
    \label{Semi-prime H-numbers}
    \begin{algorithmic}
        \FOR{all H-primes $i$}
            \FOR{all H-primes $j \ge i$}
                \STATE $i \cdot j$ is semi-prime
            \ENDFOR
        \ENDFOR
    \end{algorithmic}
\end{algorithm}

\noindent{What} remains to be done is to count the semi-prime H-numbers. After
doing this, one can get the answer from the counting array.

The complexity of our naive H-prime sieve (algorithm \ref{lst:hsieve}) is
$O(N^2)$. To calculate the semi-prime H-numbers we need to go through all
H-primes, which is $O(P^2)$ where P is the number of primes. And lastly, to
count the H-primes is linear, $O(N)$. The complexity of this entire algorithm
is therefor $O(N^2 + P^2 + N) = O(N^2)$

% TODO fix links everywhere, but how?

\end{document}
